Inference of seasonal forcings on reproduction numbers during epidemics

### 2020-04-18 PaperDraft.Rmd
some formatting in reprot

Amidst the current COVID-19 epidemic, the prediction of its future spread and the length of the epidemic becomes increasingly significant. We investigate the effects of temperature and humidity on the reproductive rate R of COVID-19, as warmer climates in the summer have historically resulted in diminished spread of influenza. Early epidemiology research in COVID-19 have used simple linear models such as ordinary least squares regression to estimate the relationship between R and climate. We estimate the reproductive rate of the virus using a novel Bayesian inference model with MCMC sampling to daily case data of various climate regions. While results are largely inconclusive, they provide us with a weak posterior update that R decreases with temperature and relative humidity.

@@

Due to inadequate public data to compute the serial interval, we use the estimated serial interval of $4.7$ days with standard deviation $2.9$ days obtained from Nishiura et al. (2020) [1]. The estimation of the latter quantity is important for policy, because infectious disease outbreaks will switch from an exponential growth regime to an exponential decay regime when the reproductive number falls below 1. The goal of mitigation or suppression policies is therefore to effect change on the reproductive number R. Many infectious diseases are believed to be seasonal, including most respiratory viruses, like the SARS-CoV-2 virus implicated in the 2019–2020 coronavirus pandemic. Wang et al. (2020) [3] have analysed how the weather of locations within China correlates with local point estimates of the reproductive number R. However, because their analysis is frequentist in nature, their estimates of the seasonal forcing do not come with any estimate of uncertainty.

@@

The baseline model we build estimates the reproductive rate using a parametric SI model, where we specify the mean and standard deviation of the serial interval. Because we do not have access to line-list SI data for COVID-19, we are unable to effectively estimate the serial interval from data, thus we take the findings of Nishiura et al. (2020) [1], which estimated the mean serial interval to be 4.7 and the standard deviation to be 2.4. We use the model detailed in Wallinga and Teunis (2004) [8] for estimating R from bayesian inference on a fixed serial interval distribution and time series data, which can be easily implemented using the EpiEstim package produced by Cori et al. (2013) [7].

For each province, we take a 7 day rolling sum on the number of new cases each day, and feed the data into the parametric SI model, which is given distribution $w(\tau)$ From there, the model uses a likelihood based approach to estiamte R at each discrete timestep, computing the likelihood that a person at timestep $i$ is infected by a person infected at timestep $j$, and summing the likelihood for each $j$ across all its infectees to compute the $R$ value at timestep $j$. Thus, for each province, we produce a daily time series estimate for $R$ on each day. This method is outlined in detail in Wallinga and Teunis (2004)[8].

@@

- $x_1$: standardized GDP per capita
- $x_2$: standardized population density
- $x_3$: standardized air temperature
- $x_4$: standardized relative humidity
- $y$: logarithm of estimated R

@@

## Discussion
The posterior estimates of our model suggest that there is some weak evidence that with increasing temperature, the reproductive number R decreases. The effect of relative humidity on the reproductive number is much less conclusive. As we only used data from Italy in our model building, there is a limited range of weather information than we would like to have. Our model does a good job estimating the reproductive number $R$ in various provinces with high confidence level and provides valuable information for policy makers. However, with no access to line-list data to estimate the serial interval number ourselves, our estimated reproductive number R is limited by the accuracy of the serial interval number we used. The estimated $R$ values tend to have wide confidence interval in the beginning dates due to the imported cases in the beginning and the lack of testing.

@@

We would also like to look into statistical simulation models which consider intercity travel and geographic proximity. As our estimated $R$ is dependent on the serial interval number, it would also be worthwhile to see the sensitivity of our model with different values of serial interval number as Flaxman has done in [6].

>>>>

Amidst the current COVID-19 epidemic, the prediction of its future spread and the length of the epidemic becomes increasingly significant. We investigate the effects of temperature and humidity on the reproductive rate $R$ of COVID-19, as warmer climates in the summer have historically resulted in diminished spread of influenza. Early epidemiology research in COVID-19 have used simple linear models such as ordinary least squares regression to estimate the relationship between $R$ and climate. We estimate the reproductive rate of the virus using a novel Bayesian inference model with MCMC sampling to daily case data of various climate regions. While results are largely inconclusive, they provide us with a weak posterior update that $R$ decreases with temperature and relative humidity.

@@

Due to inadequate public data to compute the serial interval, we use the estimated serial interval of $4.7$ days with standard deviation $2.9$ days obtained from Nishiura et al. (2020) [1]. The estimation of the latter quantity is important for policy, because infectious disease outbreaks will switch from an exponential growth regime to an exponential decay regime when the reproductive number falls below 1. The goal of mitigation or suppression policies is therefore to effect change on the reproductive number $R$. Many infectious diseases are believed to be seasonal, including most respiratory viruses, like the SARS-CoV-2 virus implicated in the 2019–2020 coronavirus pandemic. Wang et al. (2020) [3] have analysed how the weather of locations within China correlates with local point estimates of the reproductive number $R$. However, because their analysis is frequentist in nature, their estimates of the seasonal forcing do not come with any estimate of uncertainty.

@@

The baseline model we build estimates the reproductive rate using a parametric SI model, where we specify the mean and standard deviation of the serial interval. Because we do not have access to line-list SI data for COVID-19, we are unable to effectively estimate the serial interval from data, thus we take the findings of Nishiura et al. (2020) [1], which estimated the mean serial interval to be 4.7 and the standard deviation to be 2.4. We use the model detailed in Wallinga and Teunis (2004) [8] for estimating $R$ from bayesian inference on a fixed serial interval distribution and time series data, which can be easily implemented using the EpiEstim package produced by Cori et al. (2013) [7].

For each province, we take a 7 day rolling sum on the number of new cases each day, and feed the data into the parametric SI model, which is given distribution $w(\tau)$ From there, the model uses a likelihood based approach to estiamte $R$ at each discrete timestep, computing the likelihood that a person at timestep $i$ is infected by a person infected at timestep $j$, and summing the likelihood for each $j$ across all its infectees to compute the $R$ value at timestep $j$. Thus, for each province, we produce a daily time series estimate for $R$ on each day. This method is outlined in detail in Wallinga and Teunis (2004)[8].

@@

- $x_1$: standardized GDP per capita
- $x_2$: standardized population density
- $x_3$: standardized air temperature
- $x_4$: standardized relative humidity
- $y$: logarithm of estimated $R$

@@

## Discussion
The posterior estimates of our model suggest that there is some weak evidence that with increasing temperature, the reproductive number $R$ decreases. The effect of relative humidity on the reproductive number is much less conclusive. As we only used data from Italy in our model building, there is a limited range of weather information than we would like to have. Our model does a good job estimating the reproductive number $R$ in various provinces with high confidence level and provides valuable information for policy makers. However, with no access to line-list data to estimate the serial interval number ourselves, our estimated reproductive number $R$ is limited by the accuracy of the serial interval number we used. The estimated $R$ values tend to have wide confidence interval in the beginning dates due to the imported cases in the beginning and the lack of testing.

@@

We would also like to look into statistical simulation models which consider intercity travel and geographic proximity. As our estimated $R$ is dependent on the serial interval number, it would also be worthwhile to see the sensitivity of our model with different values of serial interval number as did Flaxman et al. (2020)[6].
